\morus: Biases in the Stream

### 2018-05-10 morusAC_02_Preliminaries.tex
Section 2

\MORUS is a family of authenticated ciphers designed by Wu and Huang \cite{MORUS}.
In this section, we provide a brief description of \MORUS and introduce the
notation for linear approximations.

@@

An instance of \MORUS is parametrized by a secret key $K$. During encryption, it takes as input a plaintext message $M$, a nonce $N$, and possibly some associated data $A$, and outputs a ciphertext $C$ together with an authentication tag $T$.

@@

Every part of this process relies on iterating the \StateUpdate{} function at the core of \MORUS. Each call to the \StateUpdate{} function is called a step.
In order to denote the internal state of the cipher at the $t$-th step, we use a superscript to denote the step number: the internal state at step $t$ is denoted by $S^t_0\|S^t_1\|S^t_2\|S^t_3\|S^t_4$, where $t$ is $-16$ before the initialization and $t=0$ after the initialization.

@@

\StateUpdate{} is composed of 5 rounds with similar operations. The additional input $m^t$ is used in rounds 2 to 5, but not in round 1. Each round uses the word-wise left circular shift operation, which divides a $q$-bit register value into 4 words of $w = q/4$ bits, and performs a left circular shift operation on each $w$-bit word. The word-wise circular shift constants $b_i$ in round $i$ are defined in \Cref{Tbl:rcon}. Left circular shifts on a whole $q$-bit register are also used. The register-wise circular shift constants are also listed in \Cref{Tbl:rcon}.
\begin{table}[!htb]
\centering
\caption{Left Circular Shift Constants of \texorpdfstring{\MORUS}{MORUS}} \label{Tbl:rcon}
\begin{tabular}{r||ccccc|ccccc} \hline
& \multicolumn{5}{c|}{Word-wise circular shift} & \multicolumn{5}{c}{Register-wise circular shift} \\
                  & \makebox[2em]{$b_0$} & \makebox[2em]{$b_1$} & \makebox[2em]{$b_2$} & \makebox[2em]{$b_3$} & \makebox[2em]{$b_4$} & \makebox[2em]{$b'_0$} & \makebox[2em]{$b'_1$} & \makebox[2em]{$b'_2$} & \makebox[2em]{$b'_3$} & \makebox[2em]{$b'_4$} \\ \hline
\MORUS[640]  &  5 & 31 &  7 & 22 & 13 & 32 &  64 &  96 &  64 & 32 \\
\MORUS[1280] & 13 & 46 & 38 &  7 &  4 & 64 & 128 & 192 & 128 & 64 \\ \hline
\end{tabular}
\end{table}

We denote those register-wise and word-wise left circular shift by $x \lll y$ and $x \lll_W y$, respectively. (Those are denoted by \texttt{<<<} and \texttt{Rotl\_xxx\_yy} respectively in the \MORUS design document.)

Let `$\cdot$' denote the bit-wise AND operation. $S^{t+1} \leftarrow \StateUpdate{}(S^t,m^t)$ is defined as follows.
\begin{align*}
\textrm{Round 1:} &&
S^{t+1}_0 &\leftarrow ( S^t_0 \oplus (S^t_1 \cdot S^t_2) \oplus S^t_3 ) \lll_W b_0 &
S^t_3 \leftarrow S^t_3 \lll b'_0.\\
\textrm{Round 2:} &&
S^{t+1}_1 &\leftarrow ( S^t_1 \oplus (S^t_2 \cdot S^t_3) \oplus S^t_4 \oplus m_i ) \lll_W b_1 &
S^t_4 \leftarrow S^t_4 \lll b'_1.\\
\textrm{Round 3:} &&
S^{t+1}_2 &\leftarrow ( S^t_2 \oplus (S^t_3 \cdot S^t_4) \oplus S^t_0 \oplus m_i ) \lll_W b_2 &
S^t_0 \leftarrow S^t_0 \lll b'_2.\\
\textrm{Round 4:} &&
S^{t+1}_3 &\leftarrow ( S^t_3 \oplus (S^t_4 \cdot S^t_0) \oplus S^t_1 \oplus m_i ) \lll_W b_3 &
S^t_1 \leftarrow S^t_1 \lll b'_3.\\
\textrm{Round 5:} &&
S^{t+1}_4 &\leftarrow ( S^t_4 \oplus (S^t_0 \cdot S^t_1) \oplus S^t_2 \oplus m_i ) \lll_W b_4 &
S^t_2 \leftarrow S^t_2 \lll b'_4.
\end{align*}

\subsubsection{Initialization.}
The initialization of \MORUS[640] consists of loading a 128-bit key $K_{128}$ and a 128-bit nonce $IV_{128}$ into the state, and running \StateUpdate{} for 16 steps. The key and nonce are loaded into the state as follows:
\begin{align*}
S^{-16}_0 = \textrm{IV}_{128}, &&
S^{-16}_1 = K_{128}, &&
S^{-16}_2 = 1^{128}, &&
S^{-16}_3 = const_0, &&
S^{-16}_4 = const_1.
\end{align*}
The internal state is then updated 16 times with  $\StateUpdate{}(S^t,0)$ for $t=-16,-15,\ldots,-1$. Finally, the key is XORed into the state again with $S^0_1 \leftarrow S^0_1 \oplus K_{128}$.

The initialization of \MORUS[1280] is slightly different from that of \MORUS[640] due to the difference in register size and the two possible key sizes:
\begin{itemize}
  \item[-] when the key size is 128 bits, $K = K_{128} \|K_{128}$.
  \item[-] when the key size is 256 bits, $K = K_{256}$.
\end{itemize}
The state is initialized as:
\begin{align*}
S^{-16}_0 &= \textrm{IV}_{128} \| 0^{128}, &
S^{-16}_1 &= K, &
S^{-16}_2 = 1^{256}, \\
S^{-16}_3 &= 0^{256}, &
S^{-16}_4 &= const_0\|const_1.
\end{align*}
After updating the state 16 times with \StateUpdate{}, the state is updated with $S^0_1 \leftarrow S^0_1 \oplus K$.

\subsubsection{Associated Data Processing.}
After initialization, the associated data $A$ is processed. For the padding, if the last associated data block is not a full block,  it is padded to 128 bits and 256 bits for \MORUS[640] and \MORUS[1280] respectively, using 0 bits. The padded full block is then used to update the state. If the length of $A$, denoted by `${\it adlen}$' is 0, the associated data processing phase is skipped.

The state is then updated with $S^{t+1} \leftarrow \StateUpdate{}(S^t, A^t)$ for $t=0,1,\ldots,u-1$ where $u = \lceil \frac{{\it adlen}}{128} \rceil$ and $u = \lceil \frac{{\it adlen}}{256} \rceil$ for \MORUS[640] and \MORUS[1280], respectively.

>>>>

\MORUS is a family of authenticated ciphers designed by Wu and Huang \cite{MORUS}.
An instance of \MORUS is parametrized by a secret key $K$. During encryption, it takes as input a plaintext message $M$, a nonce $N$, and possibly some associated data $A$, and outputs a ciphertext $C$ together with an authentication tag $T$.
In this section, we provide a brief description of \MORUS and introduce the
notation for linear approximations.

@@

@@

Every part of this process relies on iterating the \StateUpdate{} function at the core of \MORUS. Each call to the \StateUpdate{} function is called a step.
%In order to denote the internal state of the cipher at the $t$-th step, we use a superscript to denote the step number:
The internal state at step $t$ is denoted by $S^t_0\|S^t_1\|S^t_2\|S^t_3\|S^t_4$, where $t = -16$ before the initialization and $t=0$ after the initialization.

@@

\StateUpdate{} is composed of 5 rounds with similar operations.
The additional input $m^t$ is used in rounds 2 to 5, but not in round 1.
Each round uses the word-wise rotation (left circular shift) operation,
denoted $\lll_w$ in the following and \texttt{Rotl\_xxx\_yy} in the design document.
It divides a $q$-bit register value into 4 words of $w = q/4$ bits, and performs a rotation on each $w$-bit word.
The word-wise rotation constants $b_i$ for round $i$ are defined in \Cref{Tbl:rcon}.
Additionally, each round uses rotations on a whole $q$-bit register by a multiple of the word size,
denoted $\lll$ in the following and \texttt{<<<} in the design document.
The register-wise rotation constants $b_i'$ are also listed in \Cref{Tbl:rcon}.
%\todo{rounds $0\dots4$ or $1\dots5$?}
\begin{table}%[!bht]
  \centering
  \caption{Rotation constants $b_i$ for $\lll_w$ and $b_i'$ for $\lll$ in round $i$ of \texorpdfstring{\MORUS}{MORUS}.} \label{Tbl:rcon}
  \setlength{\tabcolsep}{6pt}
  \begin{tabular}{@{}l@{\qquad}rrrrr rrrrr@{}}
    \toprule
    & \multicolumn{5}{@{}c@{~}}{Word-wise rotation $\lll_w$} & \multicolumn{5}{@{}c@{}}{Register-wise rotation $\lll$} \\
    \cmidrule(r){2-6}
    \cmidrule(l){7-11}
    & $b_0$  & $b_1$  & $b_2$  & $b_3$  & $b_4$  
    & $b_0'$ & $b_1'$ & $b_2'$ & $b_3'$ & $b_4'$  \\
    %& \makebox[2em]{$b_0$} & \makebox[2em]{$b_1$} & \makebox[2em]{$b_2$} & \makebox[2em]{$b_3$} & \makebox[2em]{$b_4$} & \makebox[2em]{$b'_0$} & \makebox[2em]{$b'_1$} & \makebox[2em]{$b'_2$} & \makebox[2em]{$b'_3$} & \makebox[2em]{$b'_4$} \\ \hline
    \midrule
    \MORUS[640]  &  5 & 31 &  7 & 22 & 13 & 32 &  64 &  96 &  64 & 32 \\
    \MORUS[1280] & 13 & 46 & 38 &  7 &  4 & 64 & 128 & 192 & 128 & 64 \\
    \bottomrule
  \end{tabular}
\end{table}

Let `$\cdot$' denote the bit-wise AND operation. $S^{t+1} \leftarrow \StateUpdate{}(S^t,m^t)$ is defined as follows:
\begin{align*}
\textrm{Round 1:} &&
S^{t+1}_0 &\leftarrow ( S^t_0 \oplus (S^t_1 \cdot S^t_2) \oplus S^t_3 )            \lll_w b_0, &
S^t_3 \leftarrow S^t_3 \lll b'_0.\\
\textrm{Round 2:} &&
S^{t+1}_1 &\leftarrow ( S^t_1 \oplus (S^t_2 \cdot S^t_3) \oplus S^t_4 \oplus m_i ) \lll_w b_1, &
S^t_4 \leftarrow S^t_4 \lll b'_1.\\
\textrm{Round 3:} &&
S^{t+1}_2 &\leftarrow ( S^t_2 \oplus (S^t_3 \cdot S^t_4) \oplus S^t_0 \oplus m_i ) \lll_w b_2, &
S^t_0 \leftarrow S^t_0 \lll b'_2.\\
\textrm{Round 4:} &&
S^{t+1}_3 &\leftarrow ( S^t_3 \oplus (S^t_4 \cdot S^t_0) \oplus S^t_1 \oplus m_i ) \lll_w b_3, &
S^t_1 \leftarrow S^t_1 \lll b'_3.\\
\textrm{Round 5:} &&
S^{t+1}_4 &\leftarrow ( S^t_4 \oplus (S^t_0 \cdot S^t_1) \oplus S^t_2 \oplus m_i ) \lll_w b_4, &
S^t_2 \leftarrow S^t_2 \lll b'_4.
\end{align*}

\subsubsection{Initialization.}
The initialization of \MORUS[640] starts by loading the 128-bit key $K_{128}$ and the 128-bit nonce $N_{128}$ into the state together with constants $c_0, c_1$:
%and running \StateUpdate{} for 16 steps. The key and nonce are loaded into the state as follows:
\begin{align*}
S^{-16}_0 &= N_{128}, &
S^{-16}_1 &= K_{128}, &
S^{-16}_2 &= 1^{128}, &
S^{-16}_3 &= c_0, &
S^{-16}_4 &= c_1.
\end{align*}
Then, $\StateUpdate{}(S^t,0)$ is iterated 16 times for $t=-16,-15,\ldots,-1$.
Finally, the key is XORed into the state again with $S^0_1 \leftarrow S^0_1 \oplus K_{128}$.

The initialization of \MORUS[1280] differs slightly due to the difference in register size and the two possible key sizes,
and uses either
$K = K_{128} \|K_{128}$ (for \MORUS[1280-128]) or
$K = K_{256}$ (for \MORUS[1280-256])
to initialize the state:
\begin{align*}
S^{-16}_0 &= N_{128} \mathrel\| 0^{128}, &
S^{-16}_1 &= K, &
S^{-16}_2 &= 1^{256}, &
S^{-16}_3 &= 0^{256}, &
S^{-16}_4 &= c_0\mathrel\|c_1.
\end{align*}
After iterating \StateUpdate{} 16 times, the state is updated with $S^0_1 \leftarrow S^0_1 \oplus K$.

\subsubsection{Associated Data Processing.}
After initialization, the associated data $A$ is processed in blocks of $q \in \{128, 256\}$ bits.
For the padding, if the last associated data block is not a full block, it is padded to $q$ bits with zeroes.
If the length of $A$, denoted by $|A|$, is 0, then the associated data processing phase is skipped;
else, the state is updated as
\[S^{t+1} \leftarrow \StateUpdate{}(S^t, A^t) \qquad \text{for } t=0,1,\ldots,\lceil |A|/q\rceil-1.\]
